---
title: "Plant Strategies :: Chapter 4 Plant Population Demography"
author: "Daniel C. Laughlin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(popbio)
library(popdemo)
library(Rage)
library(Rcompadre)
library(RColorBrewer)
library(Ternary)
library(viridis)
library(MetBrewer)
library(dplyr)
setwd("~/OneDrive - University of Wyoming/Data/Book/github_repo")
```

# Hypothetical short-lived perennial plant Hierba r치pido
```{r}
# read in data from Table 4.1
L <- c(1,0.3,0.1,0.05,0.03,0.01,0.005)
m <- c(0,0,6,7,6,3,0)
x <- c(0:6)

# make Euler-Lotka function
euler <- function(r) sum(L * m * exp(-r * x)) - 1

# solve for r given L, m, and x
res <- uniroot(f = euler, interval = c(-100, 100), tol = 1e-8, extendInt="yes")

# r
res$root

# lambda
exp(res$root) 

```

# Make matrix population model from life table of Hierba r치pido
```{r}
# create empty matrix
A <- matrix(0,length(m)-1,length(m)-1)

# fill matrix
A[1,] <- c(m[2]*L[2]/L[1], m[3]*L[3]/L[2], m[4]*L[4]/L[3],
           m[5]*L[5]/L[4], m[6]*L[6]/L[5], m[7]*L[7]/L[6])
A[2,] <- c(L[2]/L[1],0,0,0,0,0)
A[3,] <- c(0,L[3]/L[2],0,0,0,0)
A[4,] <- c(0,0,L[4]/L[3],0,0,0)
A[5,] <- c(0,0,0,L[5]/L[4],0,0)
A[6,] <- c(0,0,0,0,L[6]/L[5],0)

# compute lambda using eigenanalysis on matrix
# this value is the same as the Euler-Lotka equation estimate
eigen(A)$values[1]

matU <- A
matU[1,] <- c(0,0,0,0,0,0)

matF <- A
matF[2:6,] <- c(0,0,0,0,0,0)

```

# Calculate life history traits from matrix model of Hierba r치pido
```{r}
# separate A into component matrices U (growth/survival) and F (fecundity)
matU <- A
matU[1,] <- c(0,0,0,0,0,0)
matF <- A
matF[2:6,] <- c(0,0,0,0,0,0)

# compute Ro net reproductive rate 
Ro <- sum(L*m)  # same as Rage::net_repro_rate(matU, matF)

# compute generation time
GenT <- log(Ro)/log(exp(res$root)) # same as Rage::gen_time(matU, matF)

# compute Keyfitz's entropy
H <- sum(-log(L)*L)/sum(L) # same as Rage::entropy_k(L)

# compute degree of iteroparity
entropy_d(L,m)

# compute r from net reproductive rate  and generation time
r <- log(Ro)/GenT

# compute age at maturity
mature_age(matU, matF) 

# compute mean life expectancy
life_expect_mean(matU)

# compute mature life expectancy
longevity(matU) 
```

# Project population of Hierba r치pido
```{r}
# set initial population sizes of each age class
Nx <- c(2,2,2,2,2,2)
12*exp(res$root)^10
project(A, vector=Nx, time=10)
projection1 <- project(A, vector=Nx, time=10, return.vec=TRUE)
projection1@vec
cols <- brewer.pal(6, name="Dark2")

par(mar=c(4,4,1,1))
matplot(x=c(0:10), projection1@vec, type="l", col=cols, lwd=3, ylim=c(0,25),
        xlab="Years (t)", ylab="Population size (N)",lty=c(1:6))
legend("top",legend=c("1","2","3","4","5","6"), horiz=TRUE,bty="n",
       lty=c(1:6), lwd=2, cex=0.7, col=cols, title="Age")
```

# Download compadre database of matrix population models
```{r}
### Download COMPADRE Rfile from https://compadre-db.org/
### see for more info: https://jonesor.github.io/Rcompadre/articles/GettingStarted.html
load("COMPADRE_v.6.21.8.0.RData")
Compadre <- as_cdb(compadre)
```

# Load Echinacea angustifolia matrix (#37 from Hurlburt's thesis, included in compadre)
```{r}
cdb_check_species(Compadre, "Echinacea angustifolia")
echang <- cdb_check_species(Compadre, "Echinacea angustifolia", return_db = TRUE)

# define A, U, and F submatrices
matA <- matA(echang)[[37]]
matU <- matU(echang)[[37]]
matF <- matF(echang)[[37]]

# name stages
classInfo <- matrixClass(echang)[[37]]
matA <- name_stages(matA, c("Seedling","Small","Medium","Large","Dormant"))
matU <- name_stages(matU, c("Seedling","Small","Medium","Large","Dormant"))
matF <- name_stages(matF, c("Seedling","Small","Medium","Large","Dormant"))

# make life table from matrix model
lifeT <- mpm_to_table(matU, matF)

# compute Keyfitz's entropy
entropy_k(lifeT$lx)

# compute degree of iteroparity
entropy_d(lifeT$lx,lifeT$mx) 

## Figure 4.4
nt <- c(220, 1245, 330, 155, 195)
projection <- project(matA, nt, time=50, return.vec=TRUE)
projection@vec
cols <- brewer.pal(5, name="Set1")

par(mar=c(4,4,1,1))
matplot(x=c(0:50), projection@vec, type="l", col=cols, lwd=3,
        xlab="", ylab="",lty=c(1:5), bty="n")
title(xlab="Years (t)", ylab="", line=2.5)
title(ylab="Population size (N)")
legend("topleft",legend=c("Seedling","Small","Medium","Large","Dormant"),
       lty=c(1:5), lwd=3, col=cols, cex=0.5, bty="n")
```

# Compute stable age distribution and reproductive value
```{r}
# use right eigenvalue to compute stable stage distribution manually
stable <- eigen(matA)$vectors[,1]/sum(eigen(matA)$vectors[,1])

# use stable.stage() function to do the same
stable.stage(matA)

# use left eigenvalue to compute reproductive value manually
eL <- (eigen(t(matA))$vectors[,1]/sum(eigen(t(matA))$vectors[,1]))
repro <- Re(eL/eL[1])

# use reproductive.value() function to do the same
reproductive.value(matA)

### compute life history traits
# net reproductive rate
net_repro_rate(matU, matF)

# generation time
gen_time(matU, matF)

# age at maturity
mature_age(matU, matF)

# longevity
longevity(matU) # mature life expectancy
```

# Sensitivity analysis
```{r}
# compute sensitivities manually
num <- (reproductive.value(matA)%*%t(stable.stage(matA))) 
den <- as.numeric(reproductive.value(matA)%*%stable.stage(matA))
sens <- (1/den)*num # solve by hand

# use sensitivity() function
sens <- sensitivity(matA)
image2(sens, mar=c(1,6,6,1), box.offset=.3, col=grey.colors(10))
```

# Elasticity analysis
```{r}
# compute elasticities manually
elas <- (matA/lambda(matA))*sens 

# use elasticity() function
elas <- elasticity(matA)

### Elasticities
image2(elas, mar=c(1,6,6,1), box.offset=.3, col=viridis_pal(1, begin=0.4, end=1)(10))

```

### Demographic triangles
```{r}
par(mfrow=c(1,2),mar=c(1,1,1,1))
TernaryPlot(lab.offset=0.15,alab="Importance of Growth (G) %",
            blab="Importance of Survival (L) %", clab="Importance of Reproduction (F) %",
            grid.lines = 2, lab.cex=0.9, axis.cex=0.8,
            grid.minor.lines = 0, main="(a)")
TernaryText(c(0.25, 0.15, 0.45), labels=c("Short-lived
herbs"), cex=0.8, font=2)
TernaryText(c(0.70, 0.15, 0.15), labels=c("Herbs of
open habitats"), cex=0.8, font=2)
TernaryText(c(0.25, 0.45, 0.15), labels=c("Forest
herbs"), cex=0.8, font=2)
TernaryText(c(0.40, 0.30, 0.20), labels=c("Shrubs"), cex=0.8, font=2)
TernaryText(c(0.1, 0.70, 0.15), labels=c("Trees"), cex=0.8, font=2)

### Demographic triangle calculations for Echinacea
stasis <- 0 + 0.114 + 0.003 + 0 + 0.161 + 0.009 + 0.459 + 0.002 + 0.005 + 0.007 + 0.017 + 0
fecundity <- sum(elas[1,2:5])
growth <- 0.044 + 0.008 + 0 + 0.046 + 0.014 + 0 + 0.003 + 0.052 + 0.01
sum(stasis, fecundity, growth)

TernaryPlot(lab.offset=0.15,alab="Importance of Growth (G) %",
            blab="Importance of Survival (L) %", clab="Importance of Reproduction (F) %",
            grid.lines = 2, lab.cex=0.9, axis.cex=0.8,
            grid.minor.lines = 0, main="(b)")
TernaryText(c(0.15, 0.60, 0.15), labels=c("Echinacea
angustifolia"), cex=0.8, font=3)
TernaryText(c(0.40, 0.35, 0.15), labels=c("Mimulus
cardinalis"), cex=0.8, font=3)
TernaryText(c(0.15, 0.15, 0.7), labels=c("Solidago
mollis"), cex=0.8, font=3)
```

# Extract mean matrix for Mimulus cardinalis in compadre
```{r}
mimcar <- cdb_check_species(Compadre, "Mimulus cardinalis", return_db = TRUE)
matA <- mean(matA(mimcar))
matU <- mean(matU(mimcar))
matF <- mean(matF(mimcar))
classInfo <- matrixClass(mimcar)[[1]]
matA <- name_stages(matA, c("Seed","Small","Large","Reproductive"))
matU <- name_stages(matU, c("Seed","Small","Large","Reproductive"))
matF <- name_stages(matF, c("Seed","Small","Large","Reproductive"))
lifeT <- mpm_to_table(matU, matF)
entropy_k(lifeT$lx)
entropy_d(lifeT$lx,lifeT$mx) # degree of iteroparity
mature_age(matU, matF) # age at maturity
life_expect_mean(matU) #mean life expectancy
longevity(matU) # mature life expectancy
net_repro_rate(matU, matF)
gen_time(matU, matF)
elas <- elasticity(matA)
vr_fecundity(matU, matF)
vr_shrinkage(matU)
vr_growth(matU)
```

# Extract mean matrix for Solidago mollis in compadre 
```{r}
solmol <- cdb_check_species(Compadre, "Solidago mollis", return_db = TRUE)
matA <- mean(matA(solmol))
matU <- mean(matU(solmol))
matF <- mean(matF(solmol))
classInfo <- matrixClass(solmol)[[1]]
matA <- name_stages(matA, c("1 yr",">= 2 yr"))
matU <- name_stages(matU, c("1 yr",">= 2 yr"))
matF <- name_stages(matF, c("1 yr",">= 2 yr"))
mature_age(matU, matF) # age at maturity
lifeT <- mpm_to_table(matU, matF)
entropy_k(lifeT$lx)
entropy_d(lifeT$lx,lifeT$mx) # degree of iteroparity
life_expect_mean(matU) #mean life expectancy
longevity(matU) # mature life expectancy
net_repro_rate(matU, matF)
gen_time(matU, matF)
net_repro_rate(matU, matF)
gen_time(matU, matF)
mature_age(matU, matF) # age at maturity
life_expect_mean(matU) #mean life expectancy
longevity(matU) # mature life expectancy
vr_fecundity(matU, matF)
vr_shrinkage(matU)
vr_growth(matU)

```

# Extract mean matrix for Silene acaulis in compadre
```{r}
silaca <- cdb_check_species(Compadre, "Silene acaulis", return_db = TRUE)
matA <- mean(matA(silaca))
matU <- mean(matU(silaca))
matF <- mean(matF(silaca))
classInfo <- matrixClass(silaca)[[1]]
matA <- name_stages(matA, classInfo$MatrixClassAuthor)
matU <- name_stages(matU, classInfo$MatrixClassAuthor)
matF <- name_stages(matF, classInfo$MatrixClassAuthor)
lifeT <- mpm_to_table(matU, matF)
entropy_k(lifeT$lx)
entropy_d(lifeT$lx,lifeT$mx)
net_repro_rate(matU, matF)
gen_time(matU, matF)
matU <- name_stages(matU, classInfo$MatrixClassAuthor)
matF <- name_stages(matF, classInfo$MatrixClassAuthor)
mature_age(matU, matF)
life_expect_mean(matU)
longevity(matU)
vr_fecundity(matU, matF)
vr_shrinkage(matU)
vr_growth(matU)
```

# Extract mean matrix for Asclepias meadii in compadre
```{r}
ascmea <- cdb_check_species(Compadre, "Asclepias meadii", return_db = TRUE)
matA <- mean(matA(ascmea))
matU <- mean(matU(ascmea))
matF <- mean(matF(ascmea))
classInfo <- matrixClass(ascmea)[[1]]
matA <- name_stages(matA, classInfo$MatrixClassAuthor)
matU <- name_stages(matU, classInfo$MatrixClassAuthor)
matF <- name_stages(matF, classInfo$MatrixClassAuthor)
lifeT <- mpm_to_table(matU, matF)
entropy_k(lifeT$lx)
entropy_d(lifeT$lx,lifeT$mx) # degree of iteroparity
net_repro_rate(matU, matF)
gen_time(matU, matF)
matU <- name_stages(matU, classInfo$MatrixClassAuthor)
matF <- name_stages(matF, classInfo$MatrixClassAuthor)
matA <- matU + matF
mature_age(matU, matF) # age at maturity
life_expect_mean(matU) #mean life expectancy
longevity(matU) # mature life expectancy
vr_fecundity(matU, matF)
vr_shrinkage(matU)
vr_growth(matU)
```

# Density dependence
```{r}
wm <- 40
N <- c(1:50)
a <- 1
w <- wm * (1 + a*N)^-1

par(mar=c(4,4,1,1))
plot(N,w, type="l",lwd=2,
     ylab="w = Mean weight per plant (g)", xlab="N = Plants per square meter")
```


# Build an Integral Projection Model based on scripts from Ellner et al. 2016, thanks to Dave Atkins for assistance with this code and data preparation
```{r}
# load data
dat <- read.csv("twoSpp.csv", header=TRUE)

# convert area to log area
dat$area_t <- log(dat$area_t)
dat$area_tplus1 <- log(dat$area_tplus1)

# subset data to analyse one species
spDF <- dat %>% filter(species=="Festuca arizonica")

# ignore individuals with area < 0.05 cm^2 generated from converting points to polygons
spDF$area_t[spDF$area_t < (log(0.05))] <- NA
spDF$area_tplus1[spDF$area_tplus1 < (log(0.05))] <- NA

## set up vector of parameters
params=data.frame(
  surv.int=NA,
  surv.slope=NA,
  #####
  growth.int=NA,
  growth.slope=NA,
  growth.sd=NA,
  #####
  flwr.int=NA,
  flwr.slope=NA,
  #####
  seed.int=NA,
  seed.slope=NA,
  #####
  recruit.size.mean=NA,
  recruit.size.sd=NA,
  #####
  establishment.prob=NA
)
```

# Fit vital rate regression models
```{r}
# 1. survival regression using logistic regression
surv.reg = glm(survives_tplus1 ~ area_t, data=spDF, family=binomial)
params$surv.int = coefficients(surv.reg)[1]
params$surv.slope = coefficients(surv.reg)[2]
summary(surv.reg)

# 2. growth regression using linear model
growth.reg=lm(area_tplus1 ~ area_t, data=spDF)
params$growth.int=coefficients(growth.reg)[1]
params$growth.slope=coefficients(growth.reg)[2]
params$growth.sd=sd(resid(growth.reg))
summary(growth.reg)

## 3. flowering probability using logistic regression
flower.reg = glm(flwr.sim ~ area_t, data=spDF, family=binomial)
params$flwr.int = coefficients(flower.reg)[1]
params$flwr.slope = coefficients(flower.reg)[2]
summary(flower.reg)

# 4. seeds regression using poisson regression
# note that the seeds in this example were simulated from an empirical relationship
seed.reg = glm(seed.sim ~ area_t, data = spDF, family = "poisson")
params$seed.int=coefficients(seed.reg)[1]
params$seed.slope=coefficients(seed.reg)[2]
summary(seed.reg)

# 5. size distribution of recruits using Gaussian distribution
params$recruit.size.mean = mean(spDF$area_t[spDF$recruit==1], na.rm =TRUE)
params$recruit.size.sd = sd(spDF$area_t[spDF$recruit==1], na.rm =TRUE)

## 6. establishment probability
params$establishment.prob = sum(spDF$recruit, na.rm = TRUE) / sum(spDF$seed.sim[spDF$flwr.sim==1], na.rm=TRUE)
```

# Plot vital rate regression models
```{r}
par(mfrow=c(2,2), mar=c(4,4,2,1))
cols <- met.brewer("Degas",7)
xx=seq(-4,8,length.out=1000) # sizes at which to evaluate predictions
plot(spDF$area_t, jitter(spDF$survives_tplus1, 0.2), xlab="log Size (t)",ylab="Survival to t+1", col=cols[4], pch=19)
lines(xx,predict(surv.reg, data.frame(area_t=xx), type="response"), col=cols[3],lwd=4)
title("(a)", adj = 0, line = 1)

plot(spDF$area_t, spDF$area_tplus1, xlab="log Size (t)", ylab="log Size (t+1)", col=cols[4], pch=19)
lines(xx,predict(growth.reg, data.frame(area_t=xx)), col=cols[3],lwd=4)
title("(b)", adj = 0, line = 1)
abline(0,1, lty=2)

plot(spDF$area_t, jitter(spDF$flwr.sim, 0.2), xlab="log Size (t)",ylab="Flowers (t)", col=cols[4], pch=19)
lines(xx,predict(flower.reg, data.frame(area_t=xx), type="response"),col=cols[3],lwd=4)
title("(c)", adj = 0, line = 1)

plot(spDF$area_t[spDF$flwr.sim==1], spDF$seed.sim[spDF$flwr.sim==1], xlab="log Size (t)", ylab="Seed Number (t)",
     col=cols[4], pch=19)
lines(xx,predict(seed.reg,data.frame(area_t=xx),type="response"),col=cols[3],lwd=4)
title("(d)", adj = 0, line = 1)
```

# Create IPM functions
```{r}
## Growth function using linear model
g_z1z <- function(area_tplus1, area_t, params)
{
  mu <- params$growth.int + params$growth.slope * area_t    # mean size next year
  sig <- params$growth.sd                                   # sd about mean
  p.den.grow <- dnorm(area_tplus1, mean = mu, sd = sig)   # pdf that you are size area_tplus1 given you were size z
  return(p.den.grow)
}

## Survival function using logistic regression
s_z <- function(area_t, params)
{
  linear.p <- params$surv.int + params$surv.slope * area_t # linear predictor
  p <- 1/(1+exp(-linear.p))                            # logistic transformation to probability
  return(p)
}

## Probability of flowering function using logistic regression
p_bz <- function(area_t, params)
{
  linear.p <- params$flwr.int + params$flwr.slope * area_t      # linear predictor
  p <- 1/(1+exp(-linear.p))                                # logistic transformation to probability
  return(p)
}

## Seed production function using poisson regression
b_z <- function(area_t, params)
{
  N <- exp(params$seed.int + params$seed.slope * area_t)    # seed production of a size z plant
  return(N)
}

## Recruit size pdf using Gaussian distribution
c_0z1 <- function(area_t, params)
{
  mu <- params$recruit.size.mean
  sig <- params$recruit.size.sd
  p.deRecr <- dnorm(area_t, mean = mu, sd = sig)        # pdf of a size z1 recruit
  return(p.deRecr)
}
```

# Define kernels
```{r}
## Define the survival kernel
P_z1z <- function (area_tplus1, area_t, params) {
  return( s_z(area_t, params) * g_z1z(area_tplus1, area_t, params) )
}

## Define the reproduction kernel
F_z1z <- function (area_tplus1, area_t, params) {
  return( p_bz(area_t, params) * b_z(area_t, params) * params$establishment.prob * c_0z1(area_tplus1, params))
}

## Build the discretized kernel
mk_K <- function(m, params, L, U) {
  # mesh points 
  h <- (U - L)/m
  meshpts <- L + ((1:m) - 1/2) * h
  P <- h * (outer(meshpts, meshpts, P_z1z, params = params))
  F <- h * (outer(meshpts, meshpts, F_z1z, params = params))
  K <- P + F
  ## compute the eigen vectors / values 
  IPM.eig.sys <- eigen(K)
  ## lambda
  lambda <- Re(IPM.eig.sys$values[1])
  return(list(lambda=lambda, K = K, meshpts = meshpts, P = P, F = F))
}

```

# Fit kernel and examine lambda
```{r}
## set upper and lower integration limits to avoid evictions
lim1 <- 0.9*min(spDF$area_t, na.rm=TRUE)
lim2 <- 1.1*max(spDF$area_t, na.rm=TRUE)

out <- mk_K(m=100, params=params, L=lim1, U=lim2)
out$lambda

```

# Plot the kernel
```{r}
par(mfrow=c(1,2), mar=c(4,4,2,1))

# model with 500 very small size classes
out <- mk_K(m=500, params=params, L=lim1, U=lim2)
image(out$meshpts, out$meshpts, t(out$K)^0.01, col = viridis_pal()(12), ylim=c(lim1,lim2),xlim=c(lim1,lim2),
      xlab="log Size (t)", ylab="log Size (t+1)", main="(a)")

# compare to a model with only 4 size classes more similar to matrix model
out <- mk_K(m=4, params=params, L=lim1, U=lim2)
image(out$meshpts, out$meshpts, t(out$K)^0.05, col = viridis_pal()(12), ylim=c(lim1,lim2),xlim=c(lim1,lim2),
      xlab="log Size (t)", ylab="log Size (t+1)", main="(b)")
```

